```
title: "Time Series Analysis: End-to-End Protocol"
author: "rsquaredata"
date: 2026-01-21
date-format: "YYYY-MM-DD"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---
```

This notebook provides an end-to-end pipeline for forecasting:

```{mermaid}
flowchart LR
    A[Data] --> B[Cleaning]
    B --> C[EDA]
    C --> D[Frequency inference]
    D --> E[Candidate models]
    E --> F[Rolling CV]
    F --> G[Model selection]
    G --> H[Refit]
    H --> I[Forecast]
    I --> J[Diagnostics]
```

# 1. Setup

```{r}
library(ggplot2)
library(ggfortify)
library(forecast)
library(fpp2)

library(readr)
library(dplyr)

library(imputeTS)
library(urca)
library(patchwork)

set.seed(123)
```

# 2. Data Ingestion

## 2.1. Load data

Expected minimal structure:  
-   `date`: Date
-   `value`: numeric

```{r}
data_raw <- read_csv("data.csv")

glimpse(data_raw)
```

## 2.2. Parse and sort dates

```{r}
data_raw <- data_raw %>%
  mutate(date = as.Date(date)) %>%
  arrange(date)

summary(data_raw)
```

# 3. Automatic Frequency and Seasonality Strategy

## 3.1. Why frequency detection matters

For daily data, `frequency = 365` is often *theoretically* correct but **statistically suboptimal** when:
-   the dominant seasonality is **weekly**
-   the sample size is limited
-   ARIMA struggles with long seasonal periods

## 3.2. Calendar regularity check

```{r}
date_diffs <- diff(data_raw$date)
table(date_diffs)
```
Interpretation:  
-   `1 day`→ daily
-   `7 days`→ weekly
-   ~30 days → monthly
-   ~90 days → quarterly

If the data is irregular (missing weekends, trading days, etc.), consider aggregating (weekly/monthly) first.

## 3.3. Frequency + seasonality strategy

`ts` supports **one main seasonal frequency**. **TBATS** can model multiple seasonal periods.

For daily data, common seasonalities:
-   weekly cycle: 7
-   yearly-ish cycle: 365 (approx)

We will build a (daily-like) `ts` object, then feed it to TBATS with multiple seasonal periods when relevant.

### 3.3.1. Choose baseline frequency for TS

This is the frequency used by `ts` only for indexing.

```{r}
# Default: daily indexing
freq_ts <- 365

# If weekly data:
# freq_ts <- 52
# If monthly data:
# freq_ts <- 12
# If quarterly data:
# freq_ts <- 4
```

### 3.3.2. Build ts object

For daily indexing: start = (year, day-of-year)

```{r}
start_year <- as.numeric(format(min(data_raw$date), "%Y"))
start_sub  <- as.numeric(format(min(data_raw$date), "%j"))

ts_data <- ts(
  data_raw$value,
  start = c(start_year, start_sub),
  frequency = freq_ts
)

ts_data
```

### 3.3.3. Automatic choice of TBATS seasonal periods

```{r}
# Automatic TBATS seasonal period selection
n <- length(ts_data)  # change to n <- length(ts_clean) after handling missing values

tbats_candidates <- list(
  weekly = c(7),
  weekly_yearly = c(7, 365)
)

# Length-based constraint
if (n < 2 * 365) {
  tbats_candidates <- list(weekly = c(7))
}

tbats_candidates
```

Below is a manual method:
```{r}
# Default: daily data often has weekly + yearly seasonalities
# tbats_seasonal_periods <- c(7, 365)

# If the data are weekly (one strong annual-ish seasonality):
# tbats_seasonal_periods <- c(52)

# If the data are monthly:
# tbats_seasonal_periods <- c(12)

# If quarterly:
# tbats_seasonal_periods <- c(4)
```
*note* : if the series is short, including `365` may be too ambitious. In that case, we only try `c(7)` or `c(12)` depending on the context.

# 4. Missing values

```{r}
ggplot_na_distribution(ts_data)
ts_clean <- na_interpolation(ts_data)
sum(is.na(ts_clean))
```

# 5. Exploratory Data Analysis (EDA)

```{r}
autoplot(ts_clean) +
  labs(title = "Time series")
```
```{r}
ggseasonplot(ts_clean) +
  labs(title = "Seasonal plot (may be messy for daily series)")
```

## 5.1. Decomposition (illustrative)

This step is used for interpretation and pedagogical purposes. It does not enforce preprocessing choices in the final pipeline.

```{r}
decomp <- decompose(ts_clean)
autoplot(decomp) +
  labs(title = "Classical decomposition (exploratory)")
```

## 5.2. Box-Jenkins exploratory analysis (pedagogical)

The following section implements a classical Box–Jenkins analysis for exploratory and pedagogical purposes.
The resulting differenced series is not used in the final model selection, which relies on cross-validation across multiple model classes.

### Candidate differencing

```{r}
x0  <- ts_clean
x1  <- diff(ts_clean, differences = 1)
xS  <- diff(ts_clean, lag = freq_ts, differences = 1)
x1S <- diff(xS, differences = 1)
```

### Visual comparison

```{r}
autoplot(x0)  + ggtitle("Original") |
autoplot(x1)  + ggtitle("Diff(1)") |
autoplot(xS)  + ggtitle("Seasonal diff") |
autoplot(x1S) + ggtitle("Diff + seasonal diff")
```

### ACF / PACF diagnostics

```{r}
par(mfrow = c(2,2))
acf(x0,  main = "ACF original")
acf(x1,  main = "ACF diff(1)")
acf(xS,  main = "ACF seasonal diff")
acf(x1S, main = "ACF diff + seasonal")
par(mfrow = c(1,1))
```

### Optional PACF

```{r}
par(mfrow = c(2,2))
pacf(x0,  main = "PACF original")
pacf(x1,  main = "PACF diff(1)")
pacf(xS,  main = "PACF seasonal diff")
pacf(x1S, main = "PACF diff + seasonal")
par(mfrow = c(1,1))
```

### Box-Jenkins ARIMA (illustrative, not selected globally)

```{r}
# Manual ARIMA inspired by ACF/PACF (illustrative)
fit_bj <- auto.arima(
  ts_clean,
  seasonal = TRUE,
  stepwise = FALSE,
  approximation = FALSE
)

checkresiduals(fit_bj)
cat("This ARIMA model is used for Box–Jenkins illustration only.\n")
```

# 6. Train/Test Split (Final Evaluation)

## 6.1. Holdout Split for Final Evaluation

We keep a final holdout test. Model selection will use rolling-origin CV inside the training set.

```{r}
h_test <- 30  # final test horizon (daily)

# weekly:  h_test <- 8
# monthly: h_test <- 12
# quarterly: h_test <- 8

train_full <- window(ts_clean, end = length(ts_clean) - h_test)
test_final <- window(ts_clean, start = length(ts_clean) - h_test + 1)

length(train_full)
length(test_final)
```

## 6.2. Exponential Smoothing Models (illustrative)

This section explicitly fits the classical exponential smoothing models (SES, Holt, Holt–Winters).

These models are included for **illustrative and pedagogical purposes**. They are **not used** for cross-validation-based model selection, which relies on broader model families (ETS, ARIMA, NNAR, TBATS).

```{r}
fit_ses  <- ses(train_full, h = h_test)
fit_holt <- holt(train_full, h = h_test)
fit_hw_add <- hw(train_full, seasonal = "additive", h = h_test)

# Optional: multiplicative if variance increases with level
# fit_hw_mul <- hw(train_full, seasonal = "multiplicative", h = h_test)
```

Quick visual comparison

```{r}
autoplot(train_full) +
  autolayer(fit_ses$mean, series = "SES") +
  autolayer(fit_holt$mean, series = "Holt") +
  autolayer(fit_hw_add$mean, series = "HW additive") +
  labs(title = "Exponential Smoothing Models (Illustrative)")
```

# 7. Rolling-Origin Cross-Validation Utilities

We will compare models using time-series CV:  
-   expanding window
-   forecasting `h_cv` steps ahead
-   compute RMSE over all folds

## RMSE

```{r}
rmse <- function(y_true, y_pred) {
  sqrt(mean((y_true - y_pred)^2, na.rm = TRUE))
}
```

## Make folds

```{r}
# Create rolling-origin folds
# initial: initial training size
# h_cv: forecasting horizon per fold
# step: shift between folds
make_folds <- function(x, initial, h_cv, step) {
  n <- length(x)
  stops <- seq(initial, n - h_cv, by = step)
  lapply(stops, function(stop_idx) {
    list(
      train = window(x, end = stop_idx),
      test  = window(x, start = stop_idx + 1, end = stop_idx + h_cv)
    )
  })
}
```

## Choose CV parameters

```{r}
# CV horizon: small (daily) so we can repeat many folds
h_cv <- 14

# For weekly/monthly, typical:
# weekly:  h_cv <- 4
# monthly: h_cv <- 6

# initial training size:
initial <- max(100, round(0.6 * length(train_full)))

# step between folds:
step <- max(1, round(0.05 * length(train_full)))

folds <- make_folds(train_full, initial = initial, h_cv = h_cv, step = step)
length(folds)
```

## TBATS seasonal-period selection via CV (Conditional)

```{r}
tbats_cv_scores <- sapply(tbats_candidates, function(sp) {
  cv_score_model("tbats", folds, seasonal_periods = sp)
})

tbats_cv_scores

tbats_seasonal_periods <- tbats_candidates[[which.min(tbats_cv_scores)]]
tbats_seasonal_periods
```

## Speed Mode (Optional)

TBATS models are computationally expensive, especially when combined with rolling-origin cross-validation and multiple seasonalities.

Speed mode provides strategies to reduce computation time while preserving statistical validity. It is recommended for exploratory analysis, long daily series, or interactive notebooks.

| **When to use speed mode**  | **When not to use speed mode** |
|-----------------------------|--------------------------------|
| • exploratory analysis</br> • long daily series (> 5–10 years)</br> • interactive experimentation | • final report</br> • grading or evaluation</br> • production model selection |

### Option 1 - Fewer folds for TBATS

```{r}
# Speed mode option 1: subsample folds for TBATS only
folds_tbats <- folds[seq(1, length(folds), by = 2)]
```
Use `folds_tbats`**only** when scoring TBATS.

### Option 2 - Conditional activation (seasonality check)

```{r}
acf_vals <- acf(ts_clean, plot = FALSE)$acf
has_strong_seasonality <- max(abs(acf_vals[-1])) > 0.3

if (!has_strong_seasonality) {
  model_names <- setdiff(model_names, "tbats")
}
```

### Option 3 - Series-length threshold

```{r}
if (length(ts_clean) > 5000) {
  model_names <- setdiff(model_names, "tbats")
}
```

# 8. Candidate Models

We include:
-   Baselines: naive, seasonal naive, mean
-   ETS
-   ARIMA
-   NNAR
-   TBATS (multi-seasonality)

## 8.1. Model Fit + Forecast Wrapper

```{r}
fit_and_forecast <- function(model_name, train, h, seasonal_periods = c(7,365)) {
  if (model_name == "naive") {
    return(naive(train, h = h))
  }
  if (model_name == "snaive") {
    return(snaive(train, h = h))
  }
  if (model_name == "meanf") {
    return(meanf(train, h = h))
  }
  if (model_name == "ets") {
    fit <- ets(train)
    return(forecast(fit, h = h))
  }
  if (model_name == "arima") {
    fit <- auto.arima(train, seasonal = TRUE, stepwise = TRUE, approximation = TRUE)
    return(forecast(fit, h = h))
  }
  if (model_name == "nnar") {
    fit <- nnetar(train)
    return(forecast(fit, h = h))
  }
  if (model_name == "tbats") {
    fit <- tbats(train, seasonal.periods = seasonal_periods)
    return(forecast(fit, h = h))
  }
  stop("Unknown model_name")
}
```

## 8.2. Cross-Validation Scoring

```{r}
model_names <- c("naive", "snaive", "meanf", "ets", "arima", "nnar", "tbats")

cv_score_model <- function(model_name, folds, seasonal_periods) {
  fold_rmses <- sapply(folds, function(fold) {
    fc <- fit_and_forecast(model_name, fold$train, h = length(fold$test), seasonal_periods)
    rmse(as.numeric(fold$test), as.numeric(fc$mean))
  })
  mean(fold_rmses, na.rm = TRUE)
}

cv_scores <- sapply(model_names, function(m) cv_score_model(m, folds, tbats_seasonal_periods))
cv_scores
```

```{r}
cv_table <- data.frame(
  model = model_names,
  cv_rmse = as.numeric(cv_scores)
) %>% arrange(cv_rmse)

cv_table

best_model_cv <- cv_table$model[1]
best_model_cv
```

# 9. Final Evaluation on Holdout Test

Now we fit the selected model on `train_full` and evaluate on `test_final`.

```{r}
fc_best <- fit_and_forecast(best_model_cv, train_full, h = length(test_final), tbats_seasonal_periods)
accuracy(fc_best, test_final)
```

We also compare to baselines on final test (sanity check):

```{r}
acc_final <- rbind(
  naive  = accuracy(naive(train_full,  h = length(test_final)), test_final)[2,],
  snaive = accuracy(snaive(train_full, h = length(test_final)), test_final)[2,],
  meanf  = accuracy(meanf(train_full,  h = length(test_final)), test_final)[2,],
  best   = accuracy(fc_best, test_final)[2,]
)

acc_final
```

## Holdout Comparison Table

```{r}
acc_holdout <- rbind(
  naive = accuracy(naive(train_full, h = h_test), test_final)[2,],
  snaive = accuracy(snaive(train_full, h = h_test), test_final)[2,],
  meanf = accuracy(meanf(train_full, h = h_test), test_final)[2,],
  ses = accuracy(fit_ses, test_final)[2,],
  holt = accuracy(fit_holt, test_final)[2,],
  hw = accuracy(fit_hw_add, test_final)[2,],
  ets = accuracy(forecast(ets(train_full), h = h_test), test_final)[2,],
  arima = accuracy(forecast(auto.arima(train_full), h = h_test), test_final)[2,],
  nnar = accuracy(forecast(nnetar(train_full), h = h_test), test_final)[2,],
  tbats = accuracy(
    forecast(tbats(train_full, seasonal.periods = tbats_seasonal_periods),
             h = h_test),
    test_final
  )[2,]
)

acc_holdout
```


# 10. Refit on Full Data + Forecast for The Future

Forecast horizon for production:

```{r}
h_future <- 30
# weekly:  h_future <- 8
# monthly: h_future <- 12
```

```{r}
final_forecast <- fit_and_forecast(best_model_cv, ts_clean, h = h_future, tbats_seasonal_periods)
final_forecast
```

# 11. Final Forecast Plot

```{r}
autoplot(final_forecast) +
  labs(
    title = "Final Forecast (V3)",
    subtitle = paste("Selected by CV:", best_model_cv,
                     "| TBATS seasonal periods:", paste(tbats_seasonal_periods, collapse = ", "))
  )
```

# 12. Diagnostics

If the selected model is ARIMA/ETS/NNAR/TBATS, `checkresiduals()` works on the fitted model object.  
But here we have a forecast object, so we refit the best model to inspect residuals properly.

```{r}
refit_best_model <- function(model_name, x, seasonal_periods) {
  if (model_name == "ets") return(ets(x))
  if (model_name == "arima") return(auto.arima(x, seasonal = TRUE))
  if (model_name == "nnar") return(nnetar(x))
  if (model_name == "tbats") return(tbats(x, seasonal.periods = seasonal_periods))
  return(NULL)  # baselines don't have meaningful residual diagnostics
}

best_fit_obj <- refit_best_model(best_model_cv, train_full, tbats_seasonal_periods)

if (!is.null(best_fit_obj)) {
  checkresiduals(best_fit_obj)
} else {
  cat("Selected model is a baseline; residual diagnostics not applicable.\n")
}
```

# 13. Methodology summary

This notebook implements a complete and reproducible workflow for time series forecasting, combining classical statistical theory with modern model selection practices.

The methodology follows these main steps:

1.	Data preparation

The raw data is parsed, sorted chronologically, and converted into a ts object. Calendar regularity is checked to ensure compatibility with time series models.

2.	Frequency and seasonality strategy

A baseline frequency is selected for indexing purposes, while potential seasonal structures are identified. For daily data, both weekly and yearly seasonalities are considered. When relevant, multi-seasonal models (TBATS) are used.

3.	Exploratory analysis

Visualization, seasonal plots, and classical decomposition are used to understand the structure of the series. A Box–Jenkins analysis (differencing, ACF/PACF) is included for pedagogical and interpretative purposes only.

4.	Illustrative classical models

Classical exponential smoothing models (SES, Holt, Holt–Winters) are fitted explicitly to illustrate the concepts taught in the course. These models are not used for final model selection.

5.	Holdout split

A final portion of the series (holdout set) is reserved and excluded from all model selection steps. This set is used only for the final evaluation of the selected model.

6.	Rolling-origin cross-validation

Model selection is performed using an expanding-window cross-validation scheme. Candidate models include baselines, ETS, ARIMA/SARIMA, NNAR, and TBATS. Performance is measured using RMSE averaged across folds.

7.	Model selection and refitting

The model with the best cross-validated performance is selected and refitted on the full training data.

8.	Final evaluation and forecasting

The selected model is evaluated on the holdout set, then refitted on the complete dataset to produce future forecasts.

9.	Diagnostics
Residual diagnostics are performed on the fitted model to assess remaining autocorrelation and model adequacy.

This approach ensures a clear separation between exploratory analysis, model selection, and final evaluation, while remaining consistent with both classical time series theory and modern validation practices.







